---
title: 'EG15: Transcriptomics analysis'
output:
  html_document:
    df_print: paged
---


```{r}
###load libraries
library("DESeq2")
library(tidyverse)
library(ggrepel)
library(dplyr)


### set output paths
figures_path <- "~/OneDrive - Charité - Universitätsmedizin Berlin/Schuelke_Lab/EG15_RNA_Seq/Brain_organoids/Figures"
```

```{r}
### Load gene(/transcript) count matrix and labels

countData <- as.matrix(read.csv("gene_count_matrix.csv", row.names="gene_id", check.names = FALSE))
colData <- read.csv("Metadata_brain_organoids.csv",
    sep=";",row.names="Sample")

colData$study_day <-as.factor(colData$study_day)
colData

### Check all sample IDs in colData are also in CountData and match their orders
all(rownames(colData) %in% colnames(countData))
countData <- countData[, rownames(colData)]
all(rownames(colData) == colnames(countData))

```
```{r}
### Create a DESeqDataSet from count matrix and labels
dds <- DESeqDataSetFromMatrix(countData = countData, 
                                colData = colData, design =  ~ study_day + Treatment + study_day:Treatment)

### pre-filtering to increase speed and reduce memory requirements
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```
```{r}
###

colData(dds)
head( counts(dds), 5 )
```

```{r}
#set the reference level
dds$Treatment <- relevel(dds$Treatment, ref = "control")

### run DESeq2 and calulate the results
dds <- DESeq(dds)
res <- results(dds)
res

```
```{r}
### Log fold change shrinkage for visualization and ranking
resultsNames(dds)

resLFC <- lfcShrink(dds, coef="Treatment_T3_pulse_vs_control", type="apeglm")
resLFC
```
```{r}
### p-values and adjusted p-values
resOrdered <- res[order(res$padj), ]
summary(res)

summary(resLFC)
```

```{r}
### How many adjusted p-values were less than 0.05?
sum(res$padj < 0.05, na.rm=TRUE)

### How many adjusted p-values were less than 0.05?
sum(resLFC$padj < 0.05, na.rm=TRUE)
```

```{r}
plotDispEsts( dds )

plotMA(resLFC, ylim=c(-10,10), alpha= 0.05, main= "MA plot with alpha=0.05")

ggsave("MA_plot.png" , device=png, dpi = 600, path = figures_path, bg = "transparent", width = 20, height = 12, units = "cm" )

```

```{r}
### plotting the genecounts
### normalizes counts by the estimated size factors (or normalization factors if these were used) and adds a pseudocount of 1/2 to allow for log scale plotting. 

plotCounts(dds, gene="ENSG00000126351.13|THRA", intgroup="study_day")
plotCounts(dds, gene="ENSG00000126351.13|THRA", intgroup=c("study_day","Treatment"))

plotCounts(dds, gene="ENSG00000111640.15|GAPDH", intgroup="study_day")
plotCounts(dds, gene="ENSG00000111640.15|GAPDH", intgroup="Treatment")
plotCounts(dds, gene="ENSG00000111640.15|GAPDH", intgroup=c("study_day","Treatment"))
```
```{r}
### exporting the results
resSig <- subset(resLFC, padj < 0.05)
resSig

write.csv(as.data.frame(resSig), 
          file="T3_vs_control_results.csv")
```

```{r}
library("pheatmap")
select <- order(rowMeans(counts(dds,normalized=TRUE)),
                decreasing=TRUE)[1:20]
df <- as.data.frame(colData(dds)[,c("study_day", "Treatment")])
ntd <- normTransform(dds)
pheatmap(assay(ntd)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=T, annotation_col=df)
```

 
 Another use of the transformed data is sample clustering. Here, we apply the dist function to the transpose of the transformed count matrix to get sample-to-sample distances.
 A heatmap of this distance matrix gives us an overview over similarities and dissimilarities between samples. We have to provide a hierarchical clustering hc to the heatmap function based on the sample distances, or else the heatmap function would calculate a clustering based on the distances between the rows/columns of the distance matrix.
 
```{r}
### clustering
vsd <- vst(dds, blind=FALSE)
sampleDists <- dist(t(assay(vsd)))

library("RColorBrewer")
library("plotly")
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- rownames(colData)
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
heatmap <-pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)

heatmap
ggsave( "Heatmap.pdf", plot=heatmap, width = 40, height = 25, units = "cm")

ggsave("Heatmap.png" ,  plot=heatmap, device=png, dpi = 600, bg = "transparent", width = 20, height = 15, units = "cm" )

```

```{r}
### PCA
plotPCA(vsd, intgroup=c("Treatment"))

plotPCA(vsd, intgroup=c("study_day"))

pcaData <- plotPCA(vsd, intgroup=c("Treatment", "study_day", "hiPSC_line.line"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData, aes(PC1, PC2, color=study_day, shape=Treatment)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()+  
   theme_light(base_size = 12)+
  geom_text_repel(aes(label = hiPSC_line.line),
                  box.padding   = 0.35, 
                  point.padding = 0.5,
                  segment.color = 'grey50',
                  max.overlaps = 100)+ 
      ggtitle("Standard PCA - top 500 genes")

ggsave("PCA_top500.png" , device=png, dpi = 600, path = figures_path, bg = "transparent", width = 20, height = 12, units = "cm" )




```



```{r}

# make PCA considering all of the genes, not only the top 500

pca <- prcomp(t(counts(dds)), scale=F) #Calculate PCs, need to transpose cause prcomp expects samples as rows and genes as columns
#plot(pca$x[,1], pca$x[,2]) #2D plot of the first 2 Principal components (PCs are contained in x of PCA) 
pca.var <- pca$sdev^2 # squared standard deviation, says how much variation in the original data each PC accounts for
pca.var.per <- round(pca.var/sum(pca.var)*100, 1) #percentages, more valuable info

barplot(pca.var.per, main="Scree Plot", xlab="Principal Component", ylab="Percent Variation") #scree plot

meta <- select(pcaData, -PC1, -PC2)

pca.data <- data.frame(sample_name=rownames(pca$x),
  PC1=pca$x[,1],
  PC2=pca$x[,2]) %>%
  right_join(meta, by = c("sample_name" = "name")) 


# plot
ggplot(pca.data, aes(PC1, PC2, color=study_day, shape=Treatment)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()+  
   theme_light(base_size = 12)+
  geom_text_repel(aes(label = hiPSC_line.line),
                  box.padding   = 0.35, 
                  point.padding = 0.5,
                  segment.color = 'grey50',
                  max.overlaps = 100)+ 
      ggtitle("Advanced PCA - all genes included")

ggsave("PCA_all_genes.png" , device=png, dpi = 600, path = figures_path, bg = "transparent", width = 20, height = 12, units = "cm" )
```


```{r}
# make screeplot
    ## calculate the variance for each gene
     rv <- rowVars(assay(vsd))

     ## select the ntop genes by variance
     select <- order(rv, decreasing=TRUE)[seq_len(min(500, length(rv)))]
     
     ## perform a PCA on the data in assay(x) for the selected genes
     pca <- prcomp(t(assay(vsd)[select,]))
     
     ## the contribution to the total variance for each component
     percentVar <- pca$sdev^2 / sum( pca$sdev^2 )*100
     
     ##plot the "percentVar"
     scree_plot=data.frame(percentVar)
     scree_plot[,2]<- c(1:16)

     colnames(scree_plot)<-c("variance","component_number")
     ggplot(scree_plot, mapping=aes(x=component_number, y=variance))+geom_bar(stat="identity")+
         ggtitle("Scree plot") +
      theme(legend.position = "right", plot.title = element_text(hjust=0.5)) + 
       ylab("% Variance")
```
```{r}
#Determine which genes have the largest effect on Variance using loading scores
loading_scores <- pca$rotation[,1] # effects due to PC1 
gene_scores <- abs(loading_scores) #so we can compare both negative and positive effects
gene_score_ranked <- sort(gene_scores, decreasing=T)
top_10_genes <- names(gene_score_ranked[1:10]) #extracts names of the top 10 genes
top_10_genes

pca$rotation[top_10_genes,1] #shows who has positive and who has negative effects 
```
 
 
 
```{r}
#Cell Death Assay using Mitochondrial RNA presence
library("biomaRt")
ensembl <- useMart("ensembl",dataset="hsapiens_gene_ensembl") #retrieves human database GRCh38.p13
filters_ensembl = listFilters(ensembl)
filters_ensembl[1:15,]
searchFilters(mart = ensembl, pattern = "MT")
listFilterValues(mart = ensembl, filter = "chromosome_name")

attributes_ensembl = listAttributes(ensembl)
attributes_ensembl[1:10,]

ensembl_mitochondrial <- getBM(mart=ensembl, attributes = c("ensembl_gene_id", "chromosome_name"), filters = "chromosome_name", values="MT")

### ENSEMBL ID need to be in the same format
counts_formatted <- mutate(as.data.frame(countData), "IDs" = rownames(countData)) %>%
  separate(IDs, ".", extra="drop") %>%
  rename("new_IDs" = ".")



#subset(m, rownames(m) %in% include_list)
mitochondrial_counts <- subset(counts_formatted, counts_formatted$new_IDs %in% ensembl_mitochondrial$ensembl_gene_id) 
 
mitochondrial_counts <- mitochondrial_counts[1:(length(mitochondrial_counts)-1)]


#plot the mitochondrial RNA counts
library(reshape2)

longData<-melt(mitochondrial_counts)
longData<-longData[longData$value!=0,]

ggplot(longData, aes(x = new_IDs, y = variable)) + 
  geom_raster(aes(fill=value)) + 
  scale_fill_gradient(low="grey90", high="red") +
  labs(y="sample", x="ENSEMBL ID", title="Mitichondrial RNA analysis") +
  theme_bw() + theme(axis.text.x=element_text(size=9, angle=90),
                     axis.text.y=element_text(size=9),
                     plot.title=element_text(size=11))



```
 
 
```{r}
### P-values distribution
## P-value histogram plot
res.df <- data.frame(Feature = rownames(res.df), res.df)
rownames(res.df) <- NULL
GO_reference <- dplyr::filter(res.df, baseMean >0) %>%
  dplyr::select(GeneID= Feature, baseMean="baseMean" ) %>%
   drop_na() %>%
  separate(GeneID, "ENSEMBL_ID", sep="\\.", remove=FALSE, extra="warn", fill="right") %>% #splits multiple IDs on same row and keeps the right "ENSEMBLID"
  separate(GeneID, c("delete","Gene"), sep="\\|", remove=FALSE, extra="drop", fill="right")%>% #splits multiple IDs on same row and keeps the right
  dplyr::select(ENSEMBL_ID)

write.csv(GO_reference, 
          file="GO_reference_list.csv", row.names=FALSE)


## Sort results by adjusted p-values
ord <- order(res.df$padj, decreasing = FALSE)
res.df <- res.df[ord, ]
res.df <- cbind(data.frame(Feature = rownames(res.df)), res.df)
rownames(res.df) <- NULL

## Adjusted p-values histogram plot
ggplot(res.df[!is.na(res.df$padj), ], aes(x = padj)) +
    geom_histogram(alpha=.5, position='identity', bins = 50) +
    labs(title=paste('Histogram of', elementMetadata(res)$description[grep('adjusted', elementMetadata(res)$description)])) +
    xlab('Adjusted p-values') +
    xlim(c(0, 1.0005))
```

BiocManager::install("ReportingTools")
citation("ReportingTools")
https://bioconductor.org/packages/devel/bioc/vignettes/ReportingTools/inst/doc/rnaseqAnalysis.pdf

```{r}
library(ReportingTools)

des2Report <- HTMLReport(shortName = 'RNAseq_analysis_with_DESeq2_P0.1',
title = 'RNA-seq analysis of differential expression: Tissues expressing high vs low THRA',
reportDirectory = "./reports")

publish(dds,des2Report, pvalueCutoff=0.1,
annotation.db="org.Mm.eg.db", factor = colData(dds)$Treatment,
reportDir="./reports")
finish(des2Report)


```


```{r}
# Volcano Plot of results
library(EnhancedVolcano)
library("biomaRt")
#ensembl <- useMart("ensembl",dataset="hsapiens_gene_ensembl") #retrieves human database GRCh38.p13
#ensembl_gene_names <- getBM(mart=ensembl, attributes = c("ensembl_gene_id", "external_gene_name"), filters = "chromosome_name", values=1:22)

#ensembl_gene_names <- ensembl_gene_names %>%
  # rename(ENSEMBL_ID=ensembl_gene_id, Gene=external_gene_name)


Volcano_list <- data.frame(GeneID=resLFC@rownames, FDR=resLFC@listData[["padj"]], log2fold_change=resLFC@listData[["log2FoldChange"]]) %>%
   drop_na() %>%
  separate(GeneID, "ENSEMBL_ID", sep="\\.", remove=FALSE, extra="warn", fill="right") %>% #splits multiple IDs on same row and keeps the right "ENSEMBLID"
   group_by(GeneID) %>%
   arrange(FDR, .by_group=T) %>% #groups by Gene name and sorts them by FDR
   top_n(-1, FDR) %>% # selects for each gene the one with lowest FDR
   distinct()


Volcano_list_genes <- Volcano_list %>%
  separate(GeneID, c("delete","Gene"), sep="\\|", remove=FALSE, extra="drop", fill="right")%>% #splits multiple IDs on same row and keeps the right
  dplyr::select(-delete)
  
Volcano_list_labels <- dplyr::filter(Volcano_list_genes, FDR<0.05)
write.csv(as.data.frame(Volcano_list_labels), 
          file="Significant_genes_results_0.05_T3_vs_control.csv")


FC <- 0.5
p <- 0.05

keyvals <- rep('grey75', nrow(Volcano_list_genes))
names(keyvals) <- rep('Not Significant', nrow(Volcano_list_genes))

#keyvals[which(abs(Volcano_list$Log2FoldChange) > FC & Volcano_list$FDR > p)] <- 'grey50'
#names(keyvals)[which(abs(Volcano_list$Log2FoldChange) > FC & Volcano_list$FDR > p)] <- 'Log2FoldChange'

#keyvals[which(abs(Volcano_list$Log2FoldChange) < FC & Volcano_list$FDR < p)] <- 'grey25'
#names(keyvals)[which(abs(Volcano_list$Log2FoldChange)  < FC & Volcano_list$FDR < p)] <- '-Log10Q'

keyvals[which(Volcano_list_genes$log2fold_change < -FC & Volcano_list_genes$FDR < p)] <- 'blue2'
names(keyvals)[which(Volcano_list_genes$log2fold_change  < -FC & Volcano_list_genes$FDR < p)] <- 'Down-regulated'

keyvals[which(Volcano_list_genes$log2fold_change > FC & Volcano_list_genes$FDR < p)] <- 'red2'
names(keyvals)[which(Volcano_list_genes$log2fold_change > FC & Volcano_list_genes$FDR < p)] <- 'Up-regulated'

unique(keyvals)
unique(names(keyvals))


volcano_plot <- EnhancedVolcano(Volcano_list_genes,
    lab = Volcano_list_genes$Gene,
    x = 'log2fold_change',
    y = 'FDR',
    #selectLab = GBA_final_list$gene ,
    #selectLab = c('ENSG00000153113','ENSG00000142192', 'ENSG00000168002', 'ENSG00000087460', 'ENSG00000148935', 'ENSG00000138162'), #for specific genes 
    pCutoff = 0.05,
    FCcutoff = 0.5,
    xlim = c(-5.5, 5.5),
    ylim = c(0, -log10(10e-12)),
    colCustom = keyvals,
    xlab = bquote(~Log[2]~ 'fold change'),
    ylab = bquote(~-Log[10] ~ italic(Padj)),
    pointSize = 2.5,
    #labSize = 3.0,
    #shape = c(6, 6, 19, 16),
    title = "Expression pattern of tissues expressing less THRA1",
    subtitle = "Differential Expression Analysis, High THRA1 expressing tissues as reference",
    caption = "Fold Change cutoff = 0.5; Padj (FDR) cutoff = 0.05",
    #legend=c('NS','Log (base 2) fold-change','P value', 'P value & Log (base 2) fold-change'),
    #legend=c('','','P value', 'P value & Log (base 2) fold-change'),
    legendPosition = "right",
    legendLabSize = 15,
    legendIconSize = 10,
    col = c("grey30", "grey30", "royalblue", "red2"),
    colAlpha = 0.8,
    drawConnectors = TRUE,
    widthConnectors = 0.6,
    colConnectors = 'grey50',
    boxedLabels=T,labSize =3)
   # hline = c(10e-8), #adds cutoff lines

volcano_plot


ggsave("Volcano_plot_T3_vs_control.png" , device=png, dpi = 600, path = figures_path, bg = "transparent", width = 35, height = 20, units = "cm" )
```

```{r}
### we want a list of the DE genes between T3 vs control for each time point
### to do this, I will subset the dds element and run DESeq
### day 26

dds_day26 <- subset(dds, select=colData(dds)$study_day=="26")

design(dds_day26) = formula(~Treatment)
dds_day26

colData(dds_day26)
head( counts(dds_day26), 5 )

#set the reference level
dds_day26$Treatment <- relevel(dds_day26$Treatment, ref = "control")


### run DESeq2 and calulate the results
dds_day26 <- DESeq(dds_day26)
res_day26 <- results(dds_day26)
res_day26

### p-values and adjusted p-values
resOrdered_day26 <- res_day26[order(res_day26$padj), ]
summary(res_day26)

### How many adjusted p-values were less than 0.05?
sum(res_day26$padj < 0.05, na.rm=TRUE)

### exporting the results
resSig_day26 <- subset(res_day26, padj < 0.05)
resSig_day26

write.csv(as.data.frame(resSig_day26), 
          file="T3_vs_control_results_day26.csv")

```



```{r}
### day 46

dds_day46 <- subset(dds, select=colData(dds)$study_day=="46")

design(dds_day46) = formula(~Treatment)
dds_day46

colData(dds_day46)
head( counts(dds_day46), 5 )

#set the reference level
dds_day46$Treatment <- relevel(dds_day46$Treatment, ref = "control")


### run DESeq2 and calulate the results
dds_day46 <- DESeq(dds_day46)
res_day46 <- results(dds_day46)
res_day46

### p-values and adjusted p-values
resOrdered_day46 <- res_day46[order(res_day46$padj), ]
summary(res_day46)

### How many adjusted p-values were less than 0.05?
sum(res_day46$padj < 0.05, na.rm=TRUE)

### exporting the results
resSig_day46 <- subset(res_day46, padj < 0.05)
resSig_day46

write.csv(as.data.frame(resSig_day46), 
          file="T3_vs_control_results_day46.csv")

```

```{r}
### day 61

dds_day61 <- subset(dds, select=colData(dds)$study_day=="61")

design(dds_day61) = formula(~Treatment)
dds_day61

colData(dds_day61)
head( counts(dds_day61), 5 )

#set the reference level
dds_day61$Treatment <- relevel(dds_day61$Treatment, ref = "control")


### run DESeq2 and calulate the results
dds_day61 <- DESeq(dds_day61)
res_day61 <- results(dds_day61)
res_day61

### p-values and adjusted p-values
resOrdered_day61 <- res_day61[order(res_day61$padj), ]
summary(res_day61)

### How many adjusted p-values were less than 0.05?
sum(res_day61$padj < 0.05, na.rm=TRUE)

### exporting the results
resSig_day61 <- subset(res_day61, padj < 0.05)
resSig_day61

write.csv(as.data.frame(resSig_day61), 
          file="T3_vs_control_results_day61.csv")

```

```{r}
### one T3 treated sample in day 26 is an outlier. I will select all the genes that are DE in both day 46 AND 61, plus the 11 genes that were significant at day 26

df_day26 <- as.data.frame(resSig_day26) %>%
  mutate("Gene"=rownames(resSig_day26)) %>%
  select(Gene,"log2FoldChange_day26"= log2FoldChange, "FDR_day26" = padj )

df_day46 <- as.data.frame(resSig_day46) %>%
  mutate("Gene"=rownames(resSig_day46)) %>%
  select(Gene,"log2FoldChange_day46"= log2FoldChange, "FDR_day46" = padj )

df_day61 <- as.data.frame(resSig_day61) %>%
  mutate("Gene"=rownames(resSig_day61)) %>%
  select(Gene,"log2FoldChange_day61"= log2FoldChange, "FDR_day61" = padj )

DE_genes <- inner_join(df_day46, df_day61, by="Gene") %>%
  #full_join(df_day26) %>%
  separate(Gene, c("ENSEMBL_ID","Gene"), sep="\\|", remove=TRUE, extra="drop", fill="right")

DE_genes

write.csv(DE_genes,  file="T3_vs_control_results_independent_noday26.csv")
```
```{r}
### Another way of getting the results from day 46 and 61, is to create a dds with only these samples and run T3 vs control

dds_day46_and_61 <- subset(dds, select=colData(dds)$study_day!="26")

design(dds_day46_and_61) = formula(~Treatment)
dds_day46_and_61

colData(dds_day46_and_61)
head( counts(dds_day46_and_61), 5 )

#set the reference level
dds_day46_and_61$Treatment <- relevel(dds_day46_and_61$Treatment, ref = "control")


### run DESeq2 and calulate the results
dds_day46_and_61 <- DESeq(dds_day46_and_61)
res_day46_and_61 <- results(dds_day46_and_61)
res_day46_and_61

### p-values and adjusted p-values
resOrdered_day46_and_61 <- res_day46_and_61[order(res_day46_and_61$padj), ]
summary(res_day46_and_61)

### How many adjusted p-values were less than 0.05?
sum(res_day46_and_61$padj < 0.05, na.rm=TRUE)

### exporting the results
resSig_day46_and_61 <- subset(res_day46_and_61, padj < 0.05)
resSig_day46_and_61

### preparing for export
df_day46_and_61 <- as.data.frame(resSig_day46_and_61) %>%
  mutate("Gene"=rownames(resSig_day46_and_61)) %>%
  select(Gene,"log2FoldChange_day46_day61"= log2FoldChange, "FDR_day46_day61" = padj ) %>%
  separate(Gene, c("ENSEMBL_ID","Gene"), sep="\\|", remove=TRUE, extra="drop", fill="right")

write.csv(df_day46_and_61, 
          file="T3_vs_control_results_day46_and_61.csv")

```

```{r}
# Volcano Plot of results of day46 and day61 together (466 genes)
library(EnhancedVolcano)

Volcano_list_day46_day61_together <- data.frame(GeneID=res_day46_and_61@rownames, FDR=res_day46_and_61@listData[["padj"]], log2fold_change=res_day46_and_61@listData[["log2FoldChange"]]) %>%
   drop_na() %>%
  separate(GeneID, c("ENSEMBL_ID","Gene"), sep="\\|", remove=TRUE, extra="drop", fill="right") %>%
   group_by(ENSEMBL_ID) %>%
   arrange(FDR, .by_group=T) %>% #groups by Gene name and sorts them by FDR
   top_n(-1, FDR) %>% # selects for each gene the one with lowest FDR
   distinct()

FC <- 0.5
p <- 0.05

keyvals <- rep('grey75', nrow(Volcano_list_day46_day61_together))
names(keyvals) <- rep('Not Significant', nrow(Volcano_list_day46_day61_together))


keyvals[which(Volcano_list_day46_day61_together$log2fold_change < -FC & Volcano_list_day46_day61_together$FDR < p)] <- 'blue2'
names(keyvals)[which(Volcano_list_day46_day61_together$log2fold_change  < -FC & Volcano_list_day46_day61_together$FDR < p)] <- 'Down-regulated'

keyvals[which(Volcano_list_day46_day61_together$log2fold_change > FC & Volcano_list_day46_day61_together$FDR < p)] <- 'red2'
names(keyvals)[which(Volcano_list_day46_day61_together$log2fold_change > FC & Volcano_list_day46_day61_together$FDR < p)] <- 'Up-regulated'

unique(keyvals)
unique(names(keyvals))


volcano_plot <- EnhancedVolcano(Volcano_list_day46_day61_together,
    lab = Volcano_list_day46_day61_together$Gene,
    x = 'log2fold_change',
    y = 'FDR',
    #selectLab = GBA_final_list$gene ,
    #selectLab = c('ENSG00000153113','ENSG00000142192', 'ENSG00000168002', 'ENSG00000087460', 'ENSG00000148935', 'ENSG00000138162'), #for specific genes 
    pCutoff = 0.05,
    FCcutoff = 0.5,
    xlim = c(-8, 8),
    ylim = c(0, -log10(10e-70)),  #check that it contains all the genes
    colCustom = keyvals,
    xlab = bquote(~Log[2]~ 'fold change'),
    ylab = bquote(~-Log[10] ~ italic(Padj)),
    pointSize = 2.5,
    #labSize = 3.0,
    #shape = c(6, 6, 19, 16),
    title = "Differentially Expressed Genes - day46 and day61 in one dds object",
    subtitle = paste0("Total of ",as.character(count(df_day46_and_61)), " significant genes"),
    caption = "Fold Change cutoff = 0.5; Padj (FDR) cutoff = 0.05",
    legendPosition = "right",
    legendLabSize = 15,
    legendIconSize = 10,
    col = c("grey30", "grey30", "royalblue", "red2"),
    colAlpha = 0.8,
    drawConnectors = TRUE,
    widthConnectors = 0.6,
    colConnectors = 'grey50',
    boxedLabels=T,labSize =3)

volcano_plot


ggsave("Volcano_plot_T3_vs_control_day46_day61_together.png" , device=png, dpi = 600, path = figures_path, bg = "transparent", width = 35, height = 20, units = "cm" )
```


```{r}
# Volcano Plot of results of day46 and day61 independently (138 genes)
library(EnhancedVolcano)

Volcano_list_day46_day61_independent <- data.frame(GeneID=res_day46@rownames, FDR=res_day46@listData[["padj"]], log2fold_change=res_day46@listData[["log2FoldChange"]]) %>%
   drop_na() %>%
  separate(GeneID, c("ENSEMBL_ID","Gene"), sep="\\|", remove=TRUE, extra="drop", fill="right") %>%
   group_by(ENSEMBL_ID) %>%
   arrange(FDR, .by_group=T) %>% #groups by Gene name and sorts them by FDR
   top_n(-1, FDR) %>% # selects for each gene the one with lowest FDR
   distinct()

FC <- 0.5
p <- 0.05

keyvals <- rep('grey75', nrow(Volcano_list_day46_day61_independent))
names(keyvals) <- rep('Not Significant', nrow(Volcano_list_day46_day61_independent))


keyvals[which(Volcano_list_day46_day61_independent$log2fold_change < -FC & Volcano_list_day46_day61_independent$FDR < p)] <- 'blue2'
names(keyvals)[which(Volcano_list_day46_day61_independent$log2fold_change  < -FC & Volcano_list_day46_day61_independent$FDR < p)] <- 'Down-regulated'

keyvals[which(Volcano_list_day46_day61_independent$log2fold_change > FC & Volcano_list_day46_day61_independent$FDR < p)] <- 'red2'
names(keyvals)[which(Volcano_list_day46_day61_independent$log2fold_change > FC & Volcano_list_day46_day61_independent$FDR < p)] <- 'Up-regulated'

unique(keyvals)
unique(names(keyvals))


volcano_plot <- EnhancedVolcano(Volcano_list_day46_day61_independent,
    lab = Volcano_list_day46_day61_independent$Gene,
    x = 'log2fold_change',
    y = 'FDR',
    #selectLab = GBA_final_list$gene ,
    #selectLab = c('ENSG00000153113','ENSG00000142192', 'ENSG00000168002', 'ENSG00000087460', 'ENSG00000148935', 'ENSG00000138162'), #for specific genes 
    pCutoff = 0.05,
    FCcutoff = 0.5,
    xlim = c(-10, 10),
    ylim = c(0, -log10(10e-120)),
    colCustom = keyvals,
    xlab = bquote(~Log[2]~ 'fold change'),
    ylab = bquote(~-Log[10] ~ italic(Padj)),
    pointSize = 2.5,
    #labSize = 3.0,
    #shape = c(6, 6, 19, 16),
    title = "Differentially Expressed Genes - day46 and day61 processed independently",
    subtitle = paste0("Total of ",as.character(count(DE_genes)), " significant genes"),
    caption = "Fold Change cutoff = 0.5; Padj (FDR) cutoff = 0.05",
    legendPosition = "right",
    legendLabSize = 15,
    legendIconSize = 10,
    col = c("grey30", "grey30", "royalblue", "red2"),
    colAlpha = 0.8,
    drawConnectors = TRUE,
    widthConnectors = 0.6,
    colConnectors = 'grey50',
    boxedLabels=T,labSize =3)

volcano_plot


ggsave("Volcano_plot_T3_vs_control_day46_day61_independent.png" , device=png, dpi = 600, path = figures_path, bg = "transparent", width = 35, height = 20, units = "cm" )
```

